\bf Baseline Drift Estimation for Time Series Data Using Quantile Trend Filtering

### 2019-03-16 Manuscript/detrendify_JCGS.tex
joe's edits 3/16

	\bigskip
	\begin{abstract}
		We address the problem of estimating smoothly varying baseline trends in time series data using quantile trend filtering. We first extend the basic framework to ensure non-crossing while estimating multiple quantile trends simultaneously. We also implement a parallelizable alternating direction method of moments (ADMM) algorithm for estimating trends in longer series. The ADMM algorthim enables the estimation of trends in a piecewise manner, both reducing the computation time and extending the limits of the method to larger data sizes. We also address smoothing parameter selection and propose a modified criterion based on the extended Bayesian Information Criterion. Through simulation studies and an application to low cost air quality sensor data, we demonstrate that our model provides better quantile trend estimates than existing methods and improves signal classification of low-cost air quality sensor output. 
	\end{abstract}

@@

In some applications, the slowly varying component $\V{\theta}$ is the signal of interest and the transient component $\V{s}$ is a vector of nuisance parameters. In other applications, the roles of $\V{\theta}$ and $\V{s}$ are reversed. The latter applies in the air quality applications motivating this work, as $\V{s}$ represents the signal of interest and $\V{\theta}$ represents a baseline drift that obscures the identification of the important transient events encoded in $\V{s}$. We describe this motivating problem further to illustrate the critical importance of being able to separate $\V{s}$ from $\V{\theta}$.

%	This work was motivated by the analysis of signals measured by low cost and portable air quality sensors, whose use has increased dramatically in the last decade. 
In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, the three sharp transient spikes at \Halley{Eric}{can you put the times at which the spikes occur here?}, but the baseline drift varies from one sensor to another obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor %using $\ell_1$-quantile-trend filtering 
 and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$ can lead to greatly improved and simplified downstream analysis. This is true for not just air quality applications, but universally across the many domains listed earlier.

@@

To tackle demixing problems, we introduce a scalable baseline estimation framework by building on \textit{$\ell_1$-trend filtering}, a relatively new nonparametric estimation framework. Our contributions are three-fold. 
\begin{itemize}
\item \cite{Kim2009} proposed using the check function as a possible extension of $\ell_1$-trend filtering but did not investigate it further. Here, we develop the basic $\ell_1$-quantile-trend-filtering framework and extend it to model multiple quantiles simultaneously with non-crossing constraints to ensure valid trend estimates. 
\item To reduce computation time and extend the method to long time series, we develop a parallelizable ADMM algorithm. 
\item Finally, we propose a modified criterion for performing model selection.
\end{itemize}

	In the rest of the paper, we detail our quantile trend filtering algorithms as well as how to choose the smoother parameter $\lambda$ (Section \ref{sec:methods}). We demonstrate through simulation studies that our proposed model provides better or comparable estimates of non-parametric quantile trends than existing methods (Section \ref{sec:simluation}). We further show that quantile trend filtering is a more effective method of drift removal for low-cost air quality sensors and results in improved signal classification compared to quantile smoothing splines (Section \ref{sec:application}). Finally, we discuss potential extensions of quantile trend filtering (Section \ref{sec:discussion}).

@@

	In this application, as well as those described in \cite{Ning2014}, \cite{marandi2015qualitative}, and \cite{pettersson2013algorithm}, the goal is to estimate the trend in the baseline not the mean. We can define the trend in the baseline as the trend in a low quantile of the data. A variety of methods for estimating quantile trends have already been proposed. \cite{Koenker1978} were the first to propose substituting the sum-of-squares term with the check function  \Eqn{check} to estimate a conditional quantile instead of the conditional mean. Later, \cite{KoenkerNgPortnoy1994} proposed quantile trend filtering with $k = 2$ producing quantile trends that are piecewise linear, but they did not consider extensions to higher order differences. Rather than using the $\ell_1$-norm to penalize the discrete differences, \cite{nychka1995nonparametric} used the smoothing spline penalty based on the square of the $\ell_2$-norm:

@@

The number of parameters to be estimated in this problem is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, all solvers will eventually break \Halley{Eric}{Can we be more specific about what we mean when we say ``break?"}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is  described in greater detail by \cite{boyd2011distributed}, but we briefly review how it can be used to iteratively solve the following equality constrained optimization problem.
\begin{equation}
\label{eq:split_objective}
\begin{split}
&\text{minimize} \; f(\V{\theta}) + g(\Vtilde{\theta}) \\
&\text{subject to} \; \M{A}\V{\theta} + \M{B}\Vtilde{\theta} = \V{c}.
\end{split}
\end{equation}
Recall that finding the minimizer to an equality constrained optimization problem is equivalent to the identifying the saddle point of the Lagrangian function associated with the problem \Eqn{split_objective}. ADMM seeks the saddle point of a related function called the augmented Lagrangian,
%The Lagrangian for the ALM problem
\begin{eqnarray*}
\mathcal{L}_{\gamma}(\V{\theta},\Vtilde{\theta},\V{\omega}) & = & f(\V{\theta}) + g(\Vtilde{\theta}) + \langle \V{\omega}, \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rangle
+ \frac{\gamma}{2} \lVert \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rVert_2^2,
\end{eqnarray*}
where the dual variable $\V{\omega}$ is a vector of Lagrange multipliers and $\gamma$ is a nonnegative tuning parameter. When $\gamma = 0$, the augmented Lagrangian coincides with the ordinary Lagrangian.

@@

	We also compare the performance of our quantile trend filtering method with three previously published methods: \texttt{npqw} which is the quantile-ll method described in \cite{Racine2017}, code was obtained from the author; \texttt{qsreg} in the \texttt{fields} R package and described in \cite{Oh2011, nychka1995nonparametric}; \texttt{rqss} available in the \texttt{quantreg} package and described in \cite{KoenkerNgPortnoy1994}.  The regularization parameter $\lambda$ for the \texttt{rqss} method is chosen using a grid search and minimizing the SIC criteria as described in \cite{KoenkerNgPortnoy1994}, the regularization parameter for \texttt{qsreg} was chosen using generalized cross-validation based on the quantile criterion \cite{Oh2011}.

@@

	To compare performance in estimating quantile trends, three simulation designs from \cite{Racine2017} were considered. For all designs $t = 1, \ldots, n$,  $x(t) = t/n$, and the response $\V{y}$ was generated as 
\begin{eqnarray*}
y(t) & = & \sin(2\pi x(t)) + \epsilon(x(t))
\end{eqnarray*}
	The three error distributions considered were 
	\begin{itemize}
		\item Gaussian: $\epsilon(x(t)) \sim N\left(0, \left(\frac{1+x(t)^2}{4}\right)^2\right)$
		\item Beta: $\epsilon(x(t)) \sim Beta(1, 11-10x(t))$
		\item Mixed normal: $\epsilon(x(t))$ is simulated from a mixture of $N(-1,1)$ and  $N(1,1)$ with mixing probability $x(t)$.
	\end{itemize}

@@

%% ----------------------------------------------------------------------
%% Peak Detection
%% ----------------------------------------------------------------------
\subsection{Peak Detection}
%	\begin{figure}[h]
%		\caption{Example of simulated peaks, baseline, and observed measurements.}
%		\includegraphics[width = \linewidth]{Figures/ex_peaks.png}
%	\end{figure}
	We use another simulation design based on our motivating application. We assume that the measured data can be represented by 
	\begin{eqnarray*}
	y(t_i) & = & \theta(t_i) + s(t_i) + \VE{\varepsilon}{i},
	\end{eqnarray*}  
	where $t_i = i$ for $i = 1, ..., n$, $\theta(t)$ is the drift component that varies smoothly over time, $s(t)$ is the true signal at time $t$, and $\VE{\varepsilon}{i}$ are i.i.d.\@ errors distributed as $N(0, 0.25^2)$. We generate $\theta(t)$ using cubic natural spline basis functions with degrees of freedom sampled from a Poisson distribution with mean parameter equal to $n/100$,  and coefficients drawn from an exponential distribution with rate 1. The true signal function $s(t)$ is assumed to be zero with peaks generated using the Gaussian density function. The number of peaks is sampled from a binomial distribution with size equal to $n$ and probability equal to $0.005$ with location parameters uniformly distributed between $1$ and $n-1$ and bandwidths uniformly distributed between $2$ and $12$. The simulated peaks were multiplied by a factor that was randomly drawn from a normal distribution with mean 20 and standard deviation of 4. One hundred datasets were generated for each $n=\{500, 1000, 2000, 4000\}$. We compare the ability of the methods to estimate the true quantiles of $y(t_i)-s(t_i)$  for $\tau \in \{0.01, 0.05, 0.1\}$ and calculate the RMSE (\Fig{peaks_rmse}). In this simulation study, our proposed method \texttt{detrend\_eBIC} method substantially outperforms the others. The \texttt{qsreg} method is comparable to the \texttt{detrend\_eBIC} method on the smaller datasets, but its performance deteriorates as the data size grows. The \texttt{npqw} and \texttt{detrend\_valid} methods both perform poorly on this design.

@@

%% ----------------------------------------------------------------------
%% Application
%% ----------------------------------------------------------------------	
\section{Application in Air Quality Monitoring}

>>>>

	\bigskip
	\begin{abstract}
		We address the problem of estimating smoothly varying baseline trends in time series data. This problem arises in several applications, including chemistry, macroeconomics, and medicine; however, our study is motivated by the analysis of data from low cost air quality sensors. Our methods extend the quantile trend filtering framework to allow the estimation of multiple quantile trends simultaneously while ensuring that the quantiles do not cross. To handle the computational challenge posed by very long time series, we propose a parallelizable alternating direction method of moments (ADMM) algorithm. The ADMM algorthim enables the estimation of trends in a piecewise manner, both reducing the computation time and extending the limits of the method to larger data sizes. We also address smoothing parameter selection and propose a modified criterion based on the extended Bayesian Information Criterion. Through simulation studies and our motivating application to low cost air quality sensor data, we demonstrate that our model provides better quantile trend estimates than existing methods and improves signal classification of low-cost air quality sensor output.
	\end{abstract}

@@

In some applications, the slowly varying component $\V{\theta}$ is the signal of interest, and the transient component $\V{s}$ is a vector of nuisance parameters. In our air quality application, the roles of $\V{\theta}$ and $\V{s}$ are reversed; $\V{s}$ represents the signal of interest and $\V{\theta}$ represents a baseline drift that obscures the identification of the important transient events encoded in $\V{s}$. We describe this motivating problem further to illustrate the critical importance of being able to separate $\V{s}$ from $\V{\theta}$.

%	This work was motivated by the analysis of signals measured by low cost and portable air quality sensors, whose use has increased dramatically in the last decade.
In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, the three sharp transient spikes at \Halley{Eric}{can you put the times at which the spikes occur here?}, but the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor %using $\ell_1$-quantile-trend filtering
 and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$ can lead to greatly improved and simplified downstream analysis. \Halley{Joe}{I would cut the next sentence, as we haven't shown this for other applications.} This is true for not just air quality applications, but universally across the many domains listed earlier.

@@

To tackle demixing problems, we introduce a scalable baseline estimation framework by building on \textit{$\ell_1$-trend filtering}, a relatively new nonparametric estimation framework. Our contributions are three-fold.
\begin{itemize}
\item \cite{Kim2009} proposed using the check function as a possible extension of $\ell_1$-trend filtering but did not investigate it further. Here, we develop the basic $\ell_1$-quantile-trend-filtering framework and extend it to model multiple quantiles simultaneously with non-crossing constraints to ensure valid trend estimates.
\item To reduce computation time and extend the method to long time series, we develop a parallelizable ADMM algorithm. \Halley{Joe}{Let's say a bit more about what this does. something like ``... a parallelizable ADMM algorithm that splits the time domain into overlapping windows and reconciles estimates from the overlapped portions.''}
\item Finally, we propose a modified criterion for performing model selection.
\end{itemize}

	In the rest of the paper, we detail our quantile trend filtering algorithms as well as how to choose the smoothing parameter (Section \ref{sec:methods}). We demonstrate through simulation studies that our proposed model provides better or comparable estimates of non-parametric quantile trends than existing methods (Section \ref{sec:simluation}). We further show that quantile trend filtering is a more effective method of drift removal for low-cost air quality sensors and results in improved signal classification compared to quantile smoothing splines (Section \ref{sec:application}). Finally, we discuss potential extensions of quantile trend filtering (Section \ref{sec:discussion}).

@@

	In this application, as well as those described in \cite{Ning2014}, \cite{marandi2015qualitative}, and \cite{pettersson2013algorithm}, the goal is to estimate the trend in the baseline not the mean. We can define the trend in the baseline as the trend in a low quantile of the data. A variety of methods for estimating quantile trends have already been proposed. \cite{Koenker1978} were the first to propose substituting the sum-of-squares term with the check function  \Eqn{check} to estimate a conditional quantile instead of the conditional mean. Later, \cite{KoenkerNgPortnoy1994} proposed quantile trend filtering with $k = 2$ producing quantile trends that are piecewise linear \Halley{Joe}{Should this say $k=1$? If $k=0$ produces piecewise constant trends, wouldn't $k=1$ produce piecewise linear trends?}, but they did not consider extensions to higher order differences. Rather than using the $\ell_1$-norm to penalize the discrete differences, \cite{nychka1995nonparametric} used the smoothing spline penalty based on the square of the $\ell_2$-norm:

@@

The number of parameters to be estimated in this problem is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, all solvers will eventually break \Halley{Eric}{Can we be more specific about what we mean when we say ``break?"}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is  described in greater detail by \cite{boyd2011distributed}, but we briefly review how it can be used to iteratively solve the following equality constrained optimization problem.
\begin{equation}
\label{eq:split_objective}
\begin{split}
&\text{minimize} \; f(\V{\theta}) + g(\Vtilde{\theta}) \\
&\text{subject to} \; \M{A}\V{\theta} + \M{B}\Vtilde{\theta} = \V{c}.
\end{split}
\end{equation}
\Eric{Joe}{I'm already lost here. Can we give some context to what $\theta$ and $\widetilde{\theta}$ are in the context of our windowing algorithm? Maybe we should say something about the windowing first before we dive head first into Lagrangians? What are $A$ and $B$? Do these simplify to something in our specific application? The stuff below might seem obvious to you, but I'm totally lost. It might be better to talk about the notation for the windows first, and how different windows will give different thetas for the overlapping parts, and then talk about equality constrained optimization as a solution for the overlapping parts.} Recall that finding the minimizer to an equality constrained optimization problem is equivalent to the identifying the saddle point of the Lagrangian function associated with the problem \Eqn{split_objective}. ADMM seeks the saddle point of a related function called the augmented Lagrangian,
%The Lagrangian for the ALM problem
\begin{eqnarray*}
\mathcal{L}_{\gamma}(\V{\theta},\Vtilde{\theta},\V{\omega}) & = & f(\V{\theta}) + g(\Vtilde{\theta}) + \langle \V{\omega}, \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rangle
+ \frac{\gamma}{2} \lVert \V{c} - \M{A}\V{\theta} - \M{B}\Vtilde{\theta} \rVert_2^2,
\end{eqnarray*}
where the dual variable $\V{\omega}$ is a vector of Lagrange multipliers and $\gamma$ is a nonnegative tuning parameter. When $\gamma = 0$, the augmented Lagrangian coincides with the ordinary Lagrangian.

@@

	We also compare the performance of our quantile trend filtering method with three previously published methods: \texttt{npqw} which is the quantile-ll method described in \cite{Racine2017}, code was obtained from the author; \texttt{qsreg} in the \texttt{fields} R package and described in \cite{Oh2011} and \cite{nychka1995nonparametric}; \texttt{rqss} available in the \texttt{quantreg} package and described in \cite{KoenkerNgPortnoy1994}.  The regularization parameter $\lambda$ for the \texttt{rqss} method is chosen using a grid search and minimizing the SIC criteria as described in \cite{KoenkerNgPortnoy1994}, the regularization parameter for \texttt{qsreg} was chosen using generalized cross-validation based on the quantile criterion \cite{Oh2011}.

@@

	To compare performance in estimating quantile trends, three simulation designs from \cite{Racine2017} were considered. For all designs $t = 1, \ldots, n$,  $x(t) = t/n$, and the response $\V{y}$ was generated as
\begin{eqnarray*}
y(t) & = & \sin(2\pi x(t)) + \epsilon(x(t))
\end{eqnarray*}
	The errors were independent and had distributions
	\begin{itemize}
		\item Gaussian: $\epsilon(x(t)) \sim N\left(0, \left(\frac{1+x(t)^2}{4}\right)^2\right)$
		\item Beta: $\epsilon(x(t)) \sim Beta(1, 11-10x(t))$
		\item Mixed normal: $\epsilon(x(t))$ is simulated from a mixture of $N(-1,1)$ and  $N(1,1)$ with mixing probability $x(t)$.
	\end{itemize}

@@

%% ----------------------------------------------------------------------
%% Peak Detection
%% ----------------------------------------------------------------------
\subsection{Peak Detection}
%	\begin{figure}[h]
%		\caption{Example of simulated peaks, baseline, and observed measurements.}
%		\includegraphics[width = \linewidth]{Figures/ex_peaks.png}
%	\end{figure}
	The second simulation design is closely motivated by our air quality analysis problem. We assume that the measured data can be represented by
	\begin{eqnarray*}
	y(t_i) & = & \theta(t_i) + s(t_i) + \VE{\varepsilon}{i},
	\end{eqnarray*}
	where $t_i = i$ for $i = 1, ..., n$, $\theta(t)$ is the drift component that varies smoothly over time, $s(t)$ is the true signal at time $t$, and $\VE{\varepsilon}{i}$ are i.i.d.\@ errors distributed as $N(0, 0.25^2)$. We generate $\theta(t)$ using cubic natural spline basis functions with degrees of freedom sampled from a Poisson distribution with mean parameter equal to $n/100$,  and coefficients drawn from an exponential distribution with rate 1. The true signal function $s(t)$ is assumed to be zero with peaks generated using the Gaussian density function. The number of peaks is sampled from a binomial distribution with size equal to $n$ and probability equal to $0.005$ with location parameters uniformly distributed between $1$ and $n-1$ and bandwidths uniformly distributed between $2$ and $12$. The simulated peaks were multiplied by a factor that was randomly drawn from a normal distribution with mean 20 and standard deviation of 4. One hundred datasets were generated for each $n=\{500, 1000, 2000, 4000\}$. \Halley{Joe}{This is a lot to take in. Can you refer to a figure that shows a simulated dataset?} We compare the ability of the methods to estimate the true quantiles of $y(t_i)-s(t_i)$  for $\tau \in \{0.01, 0.05, 0.1\}$ and calculate the RMSE (\Fig{peaks_rmse}). In this simulation study, our proposed method \texttt{detrend\_eBIC} method substantially outperforms the others. The \texttt{qsreg} method is comparable to the \texttt{detrend\_eBIC} method on the smaller datasets, but its performance deteriorates as the data size grows. The \texttt{npqw} and \texttt{detrend\_valid} methods both perform poorly on this design.

@@

%% ----------------------------------------------------------------------
%% Application
%% ----------------------------------------------------------------------
\section{Analysis of Air Quality Data}
